---
title: "viaf-normalization"
author: Alan Engel
date: "`r format(Sys.time(), '%d %B, %Y')`"
lang: en-us
output: rmarkdown::html_vignette
bibliography:
- viaf-normalization.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{viaf-normalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE}
library(viafr)
library(rnaco)
library(dplyr)
library(magrittr)
```

```{r caches, echo=FALSE}
cache <- file.path(getwd(),"Cache")
if(!dir.exists(cache)) dir.create(cache)
cache <- file.path(cache ,"Session20240707")
if(!dir.exists(cache)) dir.create(cache)
intercache <- file.path(cache,"intercache")
if(!dir.exists(intercache )) dir.create(intercache )
dnldcache <- file.path(cache,"Downloads")
if(!dir.exists(dnldcache )) dir.create(dnldcache )
```

# Name information in the form of named lists

With viafr `r packageVersion("viafr")`, name data is retrieved from viaf.xml files
in the form of named lists. This vignette works through this and functionality
for working with these named lists.

## Background

The NACO Authority Comparison Rules[@pcc2009authority][@pcc2020authority][@hickey2011naco]
call for subfields of x400 fields to be normalized individually. This requirement prompted rewriting the
function viafxml2df() and the xslt file, vc_names.xsl, to return tibbles with
named lists in the dispName column.

### Test set

The sibling package rnaco contains copies of test datasets of that the Library of Congress's
NACO CJK Funnel References Project[@naco2019naco] prepared as part of its
[CJK NACO Project](https://www.loc.gov/aba/pcc/naco/CJK.html). Fetch a set of
records that have been proofread for RDA compliance.

```{r fetchtestrecords}
rdayes <- rnaco::cjkextractja %>% dplyr::filter(rda == "Y")
ncuts <- 10
cutgap <- as.integer(nrow(rdayes)/ncuts)
testset <- rdayes %>%
        dplyr::filter(row_number() %% cutgap == 0) %>%
        dplyr::select(lccn,tag100_field,tag400_CJK,rda)
testset %>% knitr::kable(format="html")
```

The tag100_field shows the subfield structure of [MARC Field 100 entries](https://www.loc.gov/marc/bibliographic/bd100.html).

### Corresponding VIAF records

To get the corresponding VIAF records, 

```{r fn_getviafid_lccn, echo=FALSE}
getviafid_lccn <- function(lccns = NULL, cache = tempdir()) {
    library(xml2)
    library(tibble)
    library(stringi)
    dfiles <- file.path(cache,paste0(lccns,".html",sep=""))
    url <- paste0("https://viaf.org/viaf/lccn/",tolower(lccns),sep="")
    results <- curl::multi_download(url, dfiles)
    viafids <- stringi::stri_match_first_regex(results$url,pattern="https://viaf.org/viaf/(\\d++)/")[,2]
    tb <- tibble(lccn = lccns, viafID = viafids)
    return(tb)
}
```
```{r getviafid_lccn}
lcvs <- testset %>% 
	mutate(viafr::getviafid_lccn(lccn)) %>%
	select(lccn,tag100_field,viafID)
lcvs %>% knitr::kable(format="html")
```

### Fetch tag 400 data

Now fetch the tag 400 data contributed by the Library of Congress (LC). 
Just display the top few rows.

```{R fetchnames, R.options = list(width = 400)}
vnames <- viafr::cluster_names(lcvs$viafID, cache=dnldcache , intercache=intercache )
lcnames <- vnames  %>% 
    dplyr::filter(tag == '400' & source == 'LC') %>% 
    dplyr::select(viafID, normName, dispName) %>% 
    mutate(tag400_subfields = lapply(dispName, function(v) {
	ns <- names(v)
	vals <- unlist(v)
	tb <- tibble(ns,vals) %>%
		mutate(members = paste0(ns,":",vals,sep="")) %>%
		select(members) %>% unlist()
	paste(tb,collapse="<br>")
	}), .before = normName ) 
head(lcnames) %>% knitr::kable(format="html", escape=F)
```

Note that 'normName' is the normalized name as found in the VIAF xml record. 

### Join VIAF data with LC data

Join these with the lccns. Drop dispName for readability.

From left to right:

* 'lccn' is the Library of Congress control number from NACO CJK Funnel References Project's
dataset of Japanese names[@naco2019naco].
* 'tag100_field' is the personal name heading (please see the
[MARC Standard](https://www.loc.gov/marc/authority/ad100.html)). The $a and $d subfields
in MARC21 format can be observed in this column.
* 'viafID' is the VAIF cluster identifier obtained from VIAF.
* 'tag400_subfields' is the [400 - See From Tracing-Personal Name](https://www.loc.gov/marc/authority/ad400.html)
record as appears in the [VIAF cluster xml](https://viaf.org/viaf/4308165573987037800003/viaf.xml). 
Please see the [MARC Concise Formats](https://www.loc.gov/marc/MARC_2012_Concise_PDF/Part4_Authority.pdf#page=63){target="_blank"}
for details.
* 'normName' is the text from the <ns1:normalized> element of the VIAF cluster xml. This
is the form that is to be targeted in this project.


```{R joinlccns, R.options = list(width = 300)}
normdisp <- dplyr::inner_join(lcvs, lcnames, by = join_by(viafID))
normdisp %>% select(-dispName) %>% knitr::kable(format="html", escape=F)
```

Try rnaco::naco_tranform()

```{r rnaco1}
selection<-c("a","b","c","d")
rnacores <- normdisp %>% 
  mutate(rnaco = lapply(dispName, function(v){
    ns <- names(v)
    vals <-unlist(ifelse(names(v) %in% selection, rnaco::naco_transform(v,firstcomma=FALSE), v))
    tb <- tibble(ns,vals) %>%
      mutate(members = paste0(ns,":",vals,sep="")) %>%
      select(members) %>% unlist()
    paste(tb,collapse="<br>")
    }))
rnacores %>% select(lccn, tag100_field, tag400_subfields, rnaco, normName) %>% 
              knitr::kable(format="html", escape=F)
```


# References
